Enhancing Student Engagement via Reduction of Frustration with Programming Assignments using Machine Learning

### 2017-06-12 paper.tex
Some corrections
Reset column width

\author{\authorname{Mario Garcia Valdez\sup{1}, Amaury Hernandez Aguila\sup{1},
  Juan-J. Merelo\sup{2}, Pedro Castillo\sup{2} and Alejandra Mancilla Soto\sup{1}}
\affiliation{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag\)@tectijuana.edu.mx}}
\affiliation{Geneura Team and CITIC, University of Granada (Spain), 
    \email{\tt jmerelo@ugr.es}}
}

@@

Introductory programming courses are
generally regarded as difficult \cite{robins2003learning,lahtinen2005study} 
and there is a common conception that they often have a high
failure rate \cite{bennedsen2007failure}. Jenkins \cite{jenkins2001motivation} argues that there 
are multiple factors involved in this lack of success: 
\begin{itemize}
\item There is a deep relation with the expectations, attitudes, 
and previous experiences of the teaching staff and students. 
\item The nature of the subject that involves the 
learning of new abstract constructs, syntax and tools.
\item Groups are often heterogeneous and thus it is difficult to design courses 
that are beneficial for everyone. Many students begin to program when they 
are at their first year of university which means they are tackling a totally
new topic that does not respond to their habitual study approaches. 
\end{itemize}
 % Same guy here? - JJ
% same reference? You should itemize the claims - JJ
% Hope its fixed now - MG
On the other hand, Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is very
problem-solving intensive, which means that a  high precision is required because even the
slightest perturbation can render a program totally worthless. These
difficulties often frustrate students. 
Jenkins \cite{jenkins2001motivation,jenkins2002difficulty} notes that many
students expect the course to be difficult and come with the notion that they
will have to struggle, others may have a stereotyped image of a programmer,
these beliefs can negatively affect their initial motivation. During the course,
novice programmers experience many emotions, for instance frustration and
confusion when they cannot find a bug in a program, but also joy when they
successfully run a challenging program for the first time. They can also become
bored if they found the exercises too repetitive or too easy. Learning to
program is a difficult task, where emotions play a significant role. Research
has gone far to understand the role of emotion in learning, both in the process
and the outcome; for instance, in e-learning  \cite{kort2001affective,rossin2009effects}
and in programming  \cite{rodrigo2009affective,jenkins2001motivation,
bosch2013emotions,khan2007mood}.

@@

\begin{table*}[!t]
\centering
\caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
\label{tab_features}
    \begin{tabular}{ | l | l || l | l | }
    \hline
    ID          & Description           & ID               & Description  \\
    \hline
    1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
    \hline
    2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
    \hline
    3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
    \hline
    4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
    \hline
    5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
    \hline
    6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
    \hline
    7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
    \hline
    8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
    \hline
    9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
    \hline
    10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
    \hline
    11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
    \hline
    12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
    \hline
    13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
    \hline
    14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
   \hline
    15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
    \hline
    16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
    \hline
    17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
    \hline
    18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
    \hline
    19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
    \hline
    20 & Standard deviation of the previous feature. &   &  \\
    \hline
  \end{tabular}
\end{table*}

\subsubsection{Capturing the Keystroke and Mouse Data} 
While a student is trying
to solve an exercise, a script coded in JavaScript is running in the background,
which captures every keystroke, mouse movement and mouse button press. Each
capture of these events records a time-stamp in milliseconds (using the method
getTime() of JavaScript's built-in class Date) that describes when the event
occurred. If the event is a keystroke, the script captures what key was
specifically pressed, and what type of event occurred, it can be either a key-
down or a key-up event. If it is an event related to a mouse button press, the
key code of that button is recorded, as well as the type of event occurred again
key-down or key-up. Finally, if the event was a mouse movement, the mouse
coordinates inside of the web browser are recorded. The script monitors the
mouse position every 100 milliseconds, and only if the position has changed, it
records the new position. Each time a learner tries to evaluate a program, all
the data generated is sent to the server along with the code. When the result of
the execution is returned, all records are cleared and the process starts again.
There is no problem if the user leaves for a long period of time, because no
event will be triggered. If a user copies and then pastes the code from another
source, this will be recorded. There is a limitation, only the browser-editor
must be used; this could be a problem for more advanced programmers needing
specialized editors.  On the other hand a browser-based editor with the
corresponding remote execution, does not require the installation of
interpreters or compilers in learner’s computer. The code could even be written
in a mobile device or any web-enabled device. Programming exercises are evaluated using unit
tests; the results of the evaluation are shown to users. If all tests the
program is considered to be correct. The source code for the Protoboard web
based learning environment including the KD and MD functions are open source and
available as Github repositories. %at http://--------.--/----- also the code for the
sandbox is in http://--------.--/-----. % Anonimyze this, I guess - JJ

\begin{figure*}[h!t] 
\centering 
\includegraphics[width=0.9\textwidth]{editorRresult.png} 
\caption{Web-based editor used for code evaluation.}
\label{fig_editor} 
\end{figure*}

@@

A tutorial was developed to
obtain the necessary data using the proposed platform Protoboard, a web-based environment,
whose latest version has been released under an open source license. 
% can be found on-line at 
% http://------.----------.---/. 
The main functionality  used for this experiment is
shown schematically in Figure \ref{fig_process} . % Don't know what's
                                % happened to this figure
Users log in to 
Protoboard  by creating an account or by using their account credentials from the
social network Facebook. The web tutorial begins with three introductory videos that explain
the fundamentals of programming in Python, and how to use the editor to
execute their code. What follows after these videos, are 13 programming
exercises that students need to solve in sequential order. Exercises are of
incremental difficulty.  In Figure \ref{fig_editor}, 
the user interface of the programming editor is
presented.
For this experiment Protoboard was configured to allow unlimited
attempts to solve each exercise.

@@

Table \ref{tab_performance} shows the performance of each of the four classifiers is together 
with the $\kappa$ coefficients. A 10-Fold cross validation was used. 
The accuracies and $\kappa$ coefficients obtained are close to what is
usually obtained in fixed-text methods, for example, the results presented in
Epp, Lippold y Mandryk \cite{epp2011identifying}. %A reference!!! - JJ  
												 %Added	
In a fixed-text method, subjects are asked to write specific texts.											 
The results obtained with this method are satisfactory, considering that learners
were writing a program, a task comparable with free-text writing, where there
is no restriction on what has to be written. % don't get this. What free text? A program? - JJ
In the case of the $\kappa$ coefficients, it is
usual to see values below 0.2 in methods involving free-text. In this case, some
values of $\kappa$ were close to 0.5, is important to consider the values of $\kappa$
in these results because the class distribution is not uniformly distributed. As
it is observed in many works in affective recognition, decision trees normally
produce competitive accuracy. In this case the J-48 classifier obtained an
accuracy of 80.48\%, which is marginally better than the others, and an accepted $\kappa$
statistic for the kind of problem. The artificial neural network % Which one? - JJ
give the highest
accuracy in the classification of frustration.

\section{Conclusion}
The aim of this research was to evaluate the use of keyboard and mouse dynamics 
as an appropriate sensory input for an affective recognition system. The context
of use would be learning environment for programmers. In particular for learning
activities consisting in writing short programs interactively. An experimental
approach was adopted with this aim. The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and $\kappa$ below their counterparts of fixed-text, and in this
work the results obtained were similar to those works applied to fixed-text 
dynamics. A possible explanation of this would be the addition of the mouse 
dynamics features and the additional preprocessing performed on the feature vectors. 
While these are promising results further experiments are needed. Other experiments
should focus on novice programmers.
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text? 
% Is 80% effective?

Future work will focus on extracting other features from the learners'  interaction
like behavioral data, as proposed by Blikstein \cite{blikstein2011using} ; the data includes the size
of the program and changes in the code. Other authors also propose treating the use
of certain keys as features e.g., the delete key because is manly used for correcting
sentences. Also a multi-modal approach is needed.

>>>>

\author{\authorname{Mario Garcia Valdez\sup{1}, Amaury Hernandez Aguila\sup{1},
  Juan-J. Merelo\sup{2} and Alejandra Mancilla Soto\sup{1}}
\affiliation{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag\)@tectijuana.edu.mx}}
\affiliation{Geneura Team and CITIC, University of Granada (Spain), 
    \email{\tt jmerelo@ugr.es}}
}

@@

Introductory programming courses are
generally regarded as difficult \cite{robins2003learning,lahtinen2005study} 
and there is a common conception that they often have a high
failure rate \cite{bennedsen2007failure}. Jenkins \cite{jenkins2001motivation} argues that there 
are multiple factors involved in this lack of success: 
\begin{itemize}
\item There is a deep relation with the expectations, attitudes, 
and previous experiences of the teaching staff and students. 
\item The nature of the subject that involves the 
learning of new abstract constructs, syntax and tools.
\item Groups are often heterogeneous and thus it is difficult to design courses 
that are beneficial for everyone. Many students begin to program when they 
are at their first year of university which means they are tackling a totally
new topic that does not respond to their habitual study approaches. 
\end{itemize}

On the other hand, Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is very
problem-solving intensive, which means that a  high precision is required because even the
slightest perturbation can render a program totally worthless. These
difficulties often frustrate students. 
Jenkins \cite{jenkins2001motivation,jenkins2002difficulty} notes that many
students expect the course to be difficult and come with the notion that they
will have to struggle, others may have a stereotyped image of a programmer,
these beliefs can negatively affect their initial motivation. During the course,
novice programmers experience many emotions, for instance frustration and
confusion when they cannot find a bug in a program, but also joy when they
successfully run a challenging program for the first time. They can also become
bored if they found the exercises too repetitive or too easy. Learning to
program is a difficult task, where emotions play a significant role. Research
has gone far to understand the role of emotion in learning, both in the process
and the outcome; for instance, in e-learning  \cite{kort2001affective,rossin2009effects}
and in programming  \cite{rodrigo2009affective,jenkins2001motivation,
bosch2013emotions,khan2007mood}.

@@

\begin{table*}[!t]
\centering
\caption{The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
\label{tab_features}
    \begin{tabular}{ | l |p{0.4\textwidth} || l | p{0.4\textwidth} | }
    \hline
    ID          & Description           & ID               & Description  \\
    \hline
    1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
    \hline
    2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
    \hline
    3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
    \hline
    4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
    \hline
    5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
    \hline
    6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
    \hline
    7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
    \hline
    8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
    \hline
    9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
    \hline
    10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
    \hline
    11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
    \hline
    12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
    \hline
    13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
    \hline
    14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
   \hline
    15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
    \hline
    16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
    \hline
    17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
    \hline
    18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
    \hline
    19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
    \hline
    20 & Standard deviation of the previous feature. &   &  \\
    \hline
  \end{tabular}
\end{table*}

\subsubsection{Capturing the Keystroke and Mouse Data} 
While a student is trying
to solve an exercise, a script coded in JavaScript is running in the background,
which captures every keystroke, mouse movement and mouse button press. Each
capture of these events records a time-stamp in milliseconds (using the method
getTime() of JavaScript's built-in class Date) that describes when the event
occurred. If the event is a keystroke, the script captures what key was
specifically pressed, and what type of event occurred, it can be either a key-
down or a key-up event. If it is an event related to a mouse button press, the
key code of that button is recorded, as well as the type of event occurred again
key-down or key-up. Finally, if the event was a mouse movement, the mouse
coordinates inside of the web browser are recorded. The script monitors the
mouse position every 100 milliseconds, and only if the position has changed, it
records the new position. Each time a learner tries to evaluate a program, all
the data generated is sent to the server along with the code. When the result of
the execution is returned, all records are cleared and the process starts again.
There is no problem if the user leaves for a long period of time, because no
event will be triggered. If a user copies and then pastes the code from another
source, this will be recorded. There is a limitation, only the browser-editor
must be used; this could be a problem for more advanced programmers needing
specialized editors.  On the other hand a browser-based editor with the
corresponding remote execution, does not require the installation of
interpreters or compilers in learner’s computer. The code could even be written
in a mobile device or any web-enabled device. Programming exercises are evaluated using unit
tests; the results of the evaluation are shown to users. If all tests the
program is considered to be correct. The source code for the Protoboard web
based learning environment including the KD and MD functions are open source and
available as Github repositories. %at http://--------.--/----- also the code for the
sandbox is in \url{https://github.com/mariosky/sandbox}.

\begin{figure*}[h!t] 
\centering 
\includegraphics[width=0.9\textwidth]{editorRresult.png} 
\caption{Web-based editor used for code evaluation. The lower panel
  indicates (in Spanish) the tests passed successfully and the program
  output.}
\label{fig_editor} 
\end{figure*}

@@

A tutorial was developed to
obtain the necessary data using the proposed platform Protoboard, a web-based environment,
whose latest version has been released under an open source license,
and can be found online at \url{https://github.com/mariosky/protoboard}.
The main functionality  used for this experiment is
shown schematically in Figure \ref{fig_process} . % Don't know what's
                                % happened to this figure
Users log in to 
Protoboard  by creating an account or by using their account credentials from the
social network Facebook. The web tutorial begins with three introductory videos that explain
the fundamentals of programming in Python, and how to use the editor to
execute their code. What follows after these videos, are 13 programming
exercises that students need to solve in sequential order. Exercises are of
incremental difficulty.  In Figure \ref{fig_editor}, 
the user interface of the programming editor is
presented.
For this experiment Protoboard was configured to allow unlimited
attempts to solve each exercise.

@@

Table \ref{tab_performance} shows the performance of each of the four classifiers is together 
with the $\kappa$ coefficients. A 10-Fold cross validation was used. 
The accuracies and $\kappa$ coefficients obtained are close to what is
usually obtained in fixed-text methods, for example, the results presented in
Epp, Lippold y Mandryk \cite{epp2011identifying}. 	
In a fixed-text method, subjects are asked to write specific texts.											 
The results obtained with this method are satisfactory, considering that learners
were writing a program, a task comparable with free-text writing, where there
is no restriction on what has to be written. % don't get this. What free text? A program? - JJ
In the case of the $\kappa$ coefficients, it is
usual to see values below 0.2 in methods involving free-text. In this case, some
values of $\kappa$ were close to 0.5, is important to consider the values of $\kappa$
in these results because the class distribution is not uniformly distributed. As
it is observed in many works in affective recognition, decision trees normally
produce competitive accuracy. In this case the J-48 classifier obtained an
accuracy of 80.48\%, which is marginally better than the others, and an accepted $\kappa$
statistic for the kind of problem. The artificial neural network % Which one? - JJ
give the highest
accuracy in the classification of frustration.

\section{Conclusion}

The aim of this research was to evaluate the use of keyboard and mouse dynamics 
as an appropriate sensory input for an affective recognition system. The context
of use is a learning environment for programmers, in particular for learning
assignments consisting in writing short programs interactively. An experimental
approach was adopted with this objective, by effectively using the
method with real subjects using the environment.

The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and $\kappa$ measure below their counterparts of
fixed-text; however, in this
work the results obtained were similar to those works applied to fixed-text 
dynamics. A possible explanation of this would be the addition of the mouse 
dynamics features and the additional preprocessing performed on the
feature vectors.

While these are promising results further experiments are needed. Other experiments
should focus on novice programmers, but that is left as future
work. This 
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text? 
% Is 80% effective? 
future work will focus on extracting other features from the learners' interaction
such as behavioral data, as proposed by Blikstein
\cite{blikstein2011using}; the parameters proposed by this author include the size
of the program and changes in the code. Other authors also propose treating the use
of certain keys as features e.g., the delete key because is manly used for correcting
sentences. Also a multi-modal approach is needed. Since all these need
a completely new approach, a new methodology will be developed to
gather data and take measures.
